FIXME

### 2013-09-07 writeups/proposal/proposal.tex
foo
Former-commit-id: 9f687664e2b3ea19c0cd9caa81abca913524a392 [formerly 4f1d001fb1d33ee4ac474ac602e52fcbc345e3ac]
Former-commit-id: 5ec43722072896e9e7341abd4e9c9032476d948d

ACT-R \cite{Anderson2004} is a cognitive architecture that formalizes how each cognitive process of the brain (e.g., memory, learning, visual and motor) interacts to produce behavior.
The declarative memory system is a component of that architecture that models the timing, learning, and forgetting processes that make up long-term declarative memory retrieval.
The equations that make up this system can be described through a rational analysis of long-term memory retrieval.
That is, given the task of retrieving a chunk of information from long-term declarative memory,
the current context (i.e., external and internal environment state), and past experience (i.e., prior memories and exposure), 
what is the optimal behavior (i.e., the optimal chunk to retrieve from memory)?
Using Bayesian reasoning, each chunk of information in declarative memory can be assigned a prior likilihood of being retrieved again, given the prior number of times it has been needed.
Those prior probabilities for each chunk are then adjusted for the current context, so that the posterior probabilities represent the likilihood that a chunk is needed, given prior odds and current environment state.

@@

TODO: Summarize MALTA \cite{Rutledge2007}

TODO: This approach can be used to model the fan effect \cite{Rutledge2008}

TODO: However, there are open issues that need to be addressed:
TODO: Can user prior be represented in vector-based memory systems?
TODO: Can word order be represented in ACT-R?
TODO: Do they both scale?

\subsection{Incorporating Hashtag Recency}

\subsubsection{ACT-R}

TODO: ACT-R has a strong theory here.
TODO: IR literature for Twitter domain uses a model similar to ACT-R \cite{Efron2011}

\subsubsection{Vector-Based Models}

TODO: Unclear how recency will fit in with random permutation vector-based model.

>>>>

ACT-R \parencite{Anderson2004} is a cognitive architecture that formalizes how each cognitive process of the brain (e.g., memory, learning, visual and motor) interacts to produce behavior.
The declarative memory system is a component of that architecture that models the timing, learning, and forgetting processes that make up long-term declarative memory retrieval.
The equations that make up this system can be described through a rational analysis of long-term memory retrieval.
That is, given the task of retrieving a chunk of information from long-term declarative memory,
the current context (i.e., external and internal environment state), and past experience (i.e., prior memories and exposure), 
what is the optimal behavior (i.e., the optimal chunk to retrieve from memory)?
Using Bayesian reasoning, each chunk of information in declarative memory can be assigned a prior likilihood of being retrieved again, given the prior number of times it has been needed.
Those prior probabilities for each chunk are then adjusted for the current context, so that the posterior probabilities represent the likilihood that a chunk is needed, given prior odds and current environment state.

@@

Vector-based models can be thought of as a compressed representation of the full co-occurrence matrix used for the ACT-R declarative memory system.
Each model computes an activation for each word to retrieve, given prior knowledge and context.
Since both models produce activations for each chunk, one can be swapped out for another and used as the declarative memory component of the ACT-R infrastructure.
For example, \textcite{Rutledge2007} used a variant of the BEAGLE vector-based model as the DM component of ACT-R.
\textcite{Rutledge2008} showed that an ACT-R theory with a vector-based memory system can successfully model the fan effect, which is a psychological phenomina that ACT-R's current Bayesian DM system can model as well.

However, vector-based models and ACT-R's current Bayesian DM model are not identical, and there are open issues with both models that need to be addressed before further evaluating the benefits of one over the other.
For example, the frequency and recency of retrievals for each chunk are not currently used for vector-based models when calculating activation.
More recent tweets on a news site like Twitter should be more relevant, for example.
\textcite{Efron2011} tested this claim by evaluating a set of query likeihood models on returning relevant Twitter tweets from search queries.
They showed that a model incorporating recency information of the tweet produced more relevant results than models that did not pay attention to this information.

ACT-R's Bayesian model has a strong theory with respect to recency and frequency information.
It uses this information to calculate a chunk's prior activation (i.e., prior likelihood of being needed again, independent of current context).
Vector-based models compute activation based entirely on context, so chunks that resonate with context will be retrieved, irrespective to the frequency and recency of use of that chunk.
ACT-R's prior activation can also be computed for each user, so that hashtags that have been used frequently and recently for a particular user are rated with higher activation.
This type of information on a user's prior behavior is not being utilized with vector-based models currently.
It may certainly be the case that vector-based models can be modified to incorporate base-level activation, and this idea will be pursued with this research.

One strength of a vector-based model over ACT-R's Bayesian DM model is that word order can be naturally represented alongside unordered co-occerrence information.
With the ACT-R model for example, it is unclear how to represent that a particular word appeared just before a hashtag. 
It may be that multiple co-occurrence matrices must be used to represent word-order co-occurrences and unordered co-occurrences separately.
However, it would be more parsimonious if these matrices could be combined into a single matrix representation, much like how this informatino is combined for vector-based models.

\subsubsection{Activation Calculation}
